canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada's largest metropolitan areas.

### 2025-03-06 paper/Manuscript-Data-Package.Rmd
Test commit
writing data paper
Complete introduction; continue data sources description; include orcid and authors' information.

---
title: "canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada’s largest metropolitan areas."
author:
  - name: "João Pedro Figueira Amorim Parga"
    email: "joao.parga@mail.utoronto.ca"
    affiliation: CODE
    footnote: 1
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: CODE
  - name: "Robert Nutifafa Arku"
    email: "rn.arku@mail.utoronto.ca"
    affiliation: CODE
  - name: "Christopher Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliation: CODE
  - name: "Antonio Páez"
    email: "paezha@mcmaster.ca"
    affiliation: CODE        
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "In this paper, we describe the {canaccesR} package, an open data product (ODPs) created using the R statistical language. {canaccess} is a data package that provides public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas. We calculate these estimates for each Dissemination Area (DA) within these regions for the years 2019 and 2023. To do so, we use the {r5r} R package, General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data. This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility to essential services across these regions. These estimates can be used to compare different regions across Canada in terms of their accessibility and to conduct within-region equity assessments regarding access to services, which can inform improvements in transportation policies related to accessibility. The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities). Finally, as an ODPs, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."

@@

The objective of this paper is to describe the {canaccesR} Open Data Product (ODP), an open data package.
The package's main contents are public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in the years 2019 and 2023.
These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility, using the {r5r} R package [PEREIRA CITATION].
We used transport and street network, population, and employment data from different sources to arrive at our estimations.

The main contribution of this data package is to provide analysis-ready data for Canada's largest cities, thus making the field of accessibility research and urban analytics more accessible.
Estimating accessibility, i.e., the potential offered by the transportation system to reach destinations [Páez, Scott, and Morency, “Measuring Accessibility.”] requires specialized datasets and technical expertise.
By integrating and processing raw data from diverse sources, estimating TTM's for two destinations types (e.g., jobs and groceries) across Canada's largest cities of the country, and making these findings publicly available, we hope this ODP can advance the field of urban analytics in Canada.

We anticipate that these datasets will enable researchers to analyze changes in public transit accessibility to essential destinations, such as employment centers and grocery stores, in the largest urban areas of the country.
Consequently, the package addresses the need for ready-to-use public transit accessibility data, helping researchers, transit agencies, and the general public in evaluating equity in transportation accessibility within and across Canada.

@@

@@

## General Transit Feed Specification

## OpenStreetMaps

Files used for routing.

## Census Data

Employment etc.

## DMTI’s Enhanced Points of Interest

@@

@@

Below, we present some of the basic statistics of the

@@

Author 1

Author 2

Author 3

>>>>

---
title: "canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada’s largest metropolitan areas."
author:
  - name: João Pedro Figueira Amorim Parga
    email: joao.parga@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-4105-5927
    footnote: 1
  - name: Anastasia Soukhov
    email: soukhoa@mcmaster.ca
    affil-id: 2
    orcid: 0000-0003-4371-4831
  - name: Robert Nutifafa Arku
    email: rn.arku@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-2018-886X
  - name: Christopher Higgins
    email: cd.higgins@utoronto.ca
    affil-id: 3
    orcid: 0000-0002-3551-7750
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affil-id: 2
    orcid: 0000-0001-6912-9919
affiliations:
  - id: 1
    name: University of Toronto
  - id: 2
    name: McMaster University
  - id: 3
    name: University of Toronto Scarborough
# address:
#   - code: CODE
#     address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "In this paper, we describe the {canaccesR} package, an open data product (ODPs) created using the R statistical language. {canaccess} is a data package that provides public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas. We calculate these estimates for each Dissemination Area (DA) within these regions for the years 2019 and 2023. To do so, we use the {r5r} R package, General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data. This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility to essential services across these regions. These estimates can be used to compare different regions across Canada in terms of their accessibility and to conduct within-region equity assessments regarding access to services, which can inform improvements in transportation policies related to accessibility. The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities). Finally, as an ODPs, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."

@@

The objective of this paper is to describe the {canaccesR} open data package.
Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in the years 2019 and 2023.
These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility, using the {r5r} R package [@pereiraR5rRapidRealistic2021].
We used transport and street network, population, and employment data from different sources to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012] - requires specialized datasets and technical expertise.
Despite recent advancements in the field, ready-to-use available data on transportation accessibility is still sparse.
Thus, we leverage technical knowledge to make publicly available ready-to-use data on transportation accessibility.
To create the package, we integrated and processed raw data from diverse sources, estimated TTM's for two destinations types (e.g., jobs and groceries) across the largest cities in the country, and distributed these findings through this transparent and open source data product.
Our main contribution is to provide analysis-ready data for Canada's largest cities, thus making the fields of accessibility research and urban analytics more accessible.

The package's main audiences are Canadian researchers in urban planning and transportation and transportation system agencies.
We anticipate three main uses for the open data product (ODP) described in this paper.
First, the datasets allow for static assessment of the level of public transit accessibility across the country's largest cities before and after the COVID-19 pandemic.
In other words, we make it easier for those interested to compare cities and their level of public transit accessibility to essential destinations (such as employment centers and groceries stores).
Second, researchers can use the package's information as inputs to evaluate changes in accessibility throughout this period. 
Researchers may also investigate disparities in these changes across space within those cities, given the spacial character of the datasets made available here.
Third, as is common practice in transportation accessibility research, these evaluations can substantiate broader investigations on matters of justice and equity in transportation.
For example, the TTM estimates allow for evaluating the evolution of public transit's accessibility by income or spatial distribution across all Dissemination Areas (DA's) of each of the 12 cities in the sample [@pargaDemocraticAccessOur2024].

@@

The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1], which are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.

[^1]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter. We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.

@@

<!-- ## General Transit Feed Specification -->
The GTFS files from the transit agencies from all the related areas provided the information on public transit schedule and their changes from 2019 to 2023.

<!-- ## OpenStreetMaps -->
The OpenStreetMap data provided information on the transit network and was necessary for the routing used to estimate the travel time matrices.

<!-- ## Census Data -->
The 2016 and 2021 Statistics' Canada Census data provided the information on the spatial distribution of the population and the number of workplace locations, which represent the quantity of employment across the CMA's [@governmentofcanada2016CensusPopulation2016, governmentofcanada2021CensusPopulation2021].
All of the resulting travel time matrices were aggregated at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.

<!-- ## DMTI’s Enhanced Points of Interest -->
The DMTI’s Enhanced Points of Interest

@@

Other sets of data are also available at the {canaccessR} package besides the travel time matrices.
These are the boundaries, socio-economic and demographic data (e.g., population, number of dwellings, number of individuals below the Low Income Measure, etc.) of the selected CMA's, disaggregated by DA.
In addition, the package also contains aggregated population statistics (for the selected CMA's) and transit revenue and ridership data aggregated by regional and national scale.

@@

Below, we present some of the basic statistics of

@@

  - name: João Pedro Figueira Amorim Parga
    orcid: 0000-0002-4105-5927
  - name: Anastasia Soukhov
    orcid: 0000-0003-4371-4831
  - name: Robert Nutifafa Arku
    orcid: 0000-0002-2018-886X
  - name: Christopher Higgins
    orcid: 0000-0002-3551-7750
  - name: Antonio Páez
    orcid: 0000-0001-6912-9919

### 2025-03-06 paper/Manuscript-Data-Package.tex
writing data paper
Complete introduction; continue data sources description; include orcid and authors' information.

\affiliation{\affilnum{}{}}

\begin{abstract}
In this paper, we describe the \{canaccesR\} package, an open data
product (OPS) created using the R statistical language. \{canaccess\} is
a data package that provides public transit travel time estimates
(travel time matrices - TTM) to employment locations and grocery stores
across the 12 largest Canadian metropolitan areas. We calculate these
estimates for each Dissemination Area (DA) within these regions for the
years 2019 and 2023. To do so, we use the \{r5r\} R package, General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data. This data package
can be used by researchers, practitioners, and transit agencies to
estimate accessibility to essential services across these regions. These
estimates can be used to compare different regions across Canada in
terms of their accessibility and to conduct within-region equity
assessments regarding access to services, which can inform improvements
in transportation policies related to accessibility. The package is
still in its initial phase and may undergo expansions in the future by
adding TTM's for other destinations (e.g., schools, healthcare
facilities). Finally, as an OPS, the \{canaccess\} package allows for
open exploration, use, and contribution by users through its GitHub
repository.
\end{abstract}

\keywords{Accessibility; public transit; open data products; OPS; travel
time; employment; grocery stores.}

@@

The objective of this paper is to describe the \{canaccesR\} Open Data
Product (OPS), an open data package. The package's main contents are
public transit travel time matrices (TTM) estimates to employment and
groceries stores from the 12 largest Canadian metropolitan areas in the
years 2019 and 2023. These estimates were created by leveraging
expertise in data science, computer programming, and transportation
accessibility, using the \{r5r\} R package {[}PEREIRA CITATION{]}. We
used transport and street network, population, and employment data from
different sources to arrive at our estimations.

The main contribution of this data package is to provide analysis-ready
data for Canada's largest cities, thus making the field of accessibility
research and urban analytics more accessible. Estimating accessibility,
i.e., the potential offered by the transportation system to reach
destinations {[}Páez, Scott, and Morency, ``Measuring
Accessibility.''{]} requires specialized datasets and technical
expertise. By integrating and processing raw data from diverse sources,
estimating TTM's for two destinations types (e.g., jobs and groceries)
across Canada's largest cities of the country, and making these findings
publicly available, we hope this OPS can advance the field of urban
analytics in Canada.

We anticipate that these datasets will enable researchers to analyze
changes in public transit accessibility to essential destinations, such
as employment centers and grocery stores, in the largest urban areas of
the country. Consequently, the package addresses the need for
ready-to-use public transit accessibility data, helping researchers,
transit agencies, and the general public in evaluating equity in
transportation accessibility within and across Canada.

@@

@@

\subsection{General Transit Feed
Specification}\label{general-transit-feed-specification}

\subsection{OpenStreetMaps}\label{openstreetmaps}

Files used for routing.

\subsection{Census Data}\label{census-data}

Employment etc.

\subsection{DMTI's Enhanced Points of
Interest}\label{dmtis-enhanced-points-of-interest}

@@

@@

Author 1

Author 2

Author 3

>>>>

\affiliation{}

\begin{abstract}
In this paper, we describe the \{canaccesR\} package, an open data
product (ODPs) created using the R statistical language. \{canaccess\}
is a data package that provides public transit travel time estimates
(travel time matrices - TTM) to employment locations and grocery stores
across the 12 largest Canadian metropolitan areas. We calculate these
estimates for each Dissemination Area (DA) within these regions for the
years 2019 and 2023. To do so, we use the \{r5r\} R package, General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data. This data package
can be used by researchers, practitioners, and transit agencies to
estimate accessibility to essential services across these regions. These
estimates can be used to compare different regions across Canada in
terms of their accessibility and to conduct within-region equity
assessments regarding access to services, which can inform improvements
in transportation policies related to accessibility. The package is
still in its initial phase and may undergo expansions in the future by
adding TTM's for other destinations (e.g., schools, healthcare
facilities). Finally, as an ODPs, the \{canaccess\} package allows for
open exploration, use, and contribution by users through its GitHub
repository.
\end{abstract}

\keywords{Accessibility; public transit; open data products (ODPs);
travel time; employment; grocery stores.}

@@

The objective of this paper is to describe the \{canaccesR\} open data
package. Its main contents are a set of public transit travel time
matrices (TTM) estimates to employment and groceries stores from the 12
largest Canadian metropolitan areas in the years 2019 and 2023. These
estimates were created by leveraging expertise in data science, computer
programming, and transportation accessibility, using the \{r5r\} R
package \citep{pereiraR5rRapidRealistic2021}. We used transport and
street network, population, and employment data from different sources
to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation
system to reach destinations
\citep{paezMeasuringAccessibilityPositive2012} - requires specialized
datasets and technical expertise. Despite recent advancements in the
field, ready-to-use available data on transportation accessibility is
still sparse. Thus, we leverage technical knowledge to make publicly
available ready-to-use data on transportation accessibility. To create
the package, we integrated and processed raw data from diverse sources,
estimated TTM's for two destinations types (e.g., jobs and groceries)
across the largest cities in the country, and distributed these findings
through this transparent and open source data product. Our main
contribution is to provide analysis-ready data for Canada's largest
cities, thus making the fields of accessibility research and urban
analytics more accessible.

The package's main audiences are Canadian researchers in urban planning
and transportation and transportation system agencies. We anticipate
three main uses for the open data product (ODP) described in this paper.
First, the datasets allow for static assessment of the level of public
transit accessibility across the country's largest cities before and
after the COVID-19 pandemic. In other words, we make it easier for those
interested to compare cities and their level of public transit
accessibility to essential destinations (such as employment centers and
groceries stores). Second, researchers can use the package's information
as inputs to evaluate changes in accessibility throughout this period.
Researchers may also investigate disparities in these changes across
space within those cities, given the spacial character of the datasets
made available here. Third, as is common practice in transportation
accessibility research, these evaluations can substantiate broader
investigations on matters of justice and equity in transportation. For
example, the TTM estimates allow for evaluating the evolution of public
transit's accessibility by income or spatial distribution across all
Dissemination Areas (DA's) of each of the 12 cities in the sample
\citep{pargaDemocraticAccessOur2024}.

@@

The locations included in the data package comprise the 12 largest
(population-wise) Census metropolitan areas (CMA's) based on the 2021
Canadian Census \citep{governmentofcanada2021CensusPopulation2021}
\footnote{We included Oshawa, Ontario, as part of the Greater Toronto
  Area (GTA) because of the former's proximity to the latter. We also
  included Abbotsford-Mission, British Columbia, as part of the
  Vancouver metropolitan area because of the former's proximity to a
  transit station on the region's West Coast Express commuter rail line.},
which are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary,
Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo,
London, and Halifax.

@@

The GTFS files from the transit agencies from all the related areas
provided the information on public transit schedule and their changes
from 2019 to 2023.

The OpenStreetMap data provided information on the transit network and
was necessary for the routing used to estimate the travel time matrices.

The 2016 and 2021 Statistics' Canada Census data provided the
information on the spatial distribution of the population and the number
of workplace locations, which represent the quantity of employment
across the CMA's
\citep[governmentofcanada2021CensusPopulation2021]{governmentofcanada2016CensusPopulation2016}.
All of the resulting travel time matrices were aggregated at the
Dissemination Area (DA) level, which comprise the fundamental unit of
analysis in data package.

The DMTI's Enhanced Points of Interest

@@

Other sets of data are also available at the \{canaccessR\} package
besides the travel time matrices. These are the boundaries,
socio-economic and demographic data (e.g., population, number of
dwellings, number of individuals below the Low Income Measure, etc.) of
the selected CMA's, disaggregated by DA. In addition, the package also
contains aggregated population statistics (for the selected CMA's) and
transit revenue and ridership data aggregated by regional and national
scale.

@@

\begin{itemize}
\tightlist
\item
  name: João Pedro Figueira Amorim Parga orcid: 0000-0002-4105-5927
\item
  name: Anastasia Soukhov orcid: 0000-0003-4371-4831
\item
  name: Robert Nutifafa Arku orcid: 0000-0002-2018-886X
\item
  name: Christopher Higgins orcid: 0000-0002-3551-7750
\item
  name: Antonio Páez orcid: 0000-0001-6912-9919
\end{itemize}
